---
title: ""
bibliography: "misc/ref.bib"
format: 
  docx:
    reference-doc: "ref.docx"
editor_options: 
  chunk_output_type: console
execute: 
  echo: false
  warning: false
  error: false
---

**Title**: Enhancing Cycling Performance: Unveiling the Interplay of Jump Force and Neuromuscular Efficiency on Isokinetic Torque in Elite Cyclists.

**Authors**: Diego Mabe Castro^1,2^, Matías Castillo-Aguilar^1,2^, Cristian Núñez-Espinosa^2,3^\*.

::: {custom-style="Afiliation"}
^1^ Kinesiology Department, University of Magallanes, Punta Arenas, Chile.

^2^ Centro Asistencial de Docencia e Investigación (CADI-UMAG), University of Magallanes, Punta Arenas, Chile

^3^ School of Medicine, University of Magallanes, Punta Arenas, Chile.
:::

### \*Correspondence: 

Cristian Núñez-Espinosa, School of Medicine, University of Magallanes, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411.

\newpage

## Abstract 

**Objective**: [...]. **Material and methods**: [...]. **Results**: [...]. **Conclusion**: [...].

**Keywords**: [...].

\newpage

# Introduction

<!---- Antecedentes históricos ---->
[...].

<!---- Conceptualización del problema ---->
[...].

<!---- Hipótesis y objetivos ---->
[...].

# Material y methods

## Participants

[...]. 

## Instruments

### Instrument 1 

[...].

### Instrument 2

[...].

### Instrument 3 

[...].

### Instrument 4 

[...].


## Procedure

[...].

[...].  

## Statistical Analyses

### Framework

We used a Bayesian framework to explore the complex interactions among key performance factors in 11 high-performance cyclists. Bayesian models are preferred over traditional frequentist methods because they can quantify uncertainty and estimate model parameters more comprehensively. Moreover, Bayesian inference allows for the inclusion of prior knowledge about model parameters, which enables the integration of existing findings and past research. This enhances robustness, especially when dealing with limited data or complex models.

The Bayesian framework provides posterior distributions, allowing for a probabilistic interpretation of parameter estimates with credible intervals that reflect parameter uncertainty. Through Bayesian generalized linear models, we analyzed the relationships between neuromuscular efficiency through EMG recordings during FTP test and performance variables associated with jump and FTP test itself. In addition to this we also considered the inclusion of moderating variables to take into account for confounding variables in the models.

### Model-based approach

#### Neuromuscular efficiency

In the first group of models, our goal was to explore how EMG reflects neuromuscular efficiency under a high taxing assessment, and how this is influenced by key performance indicators such as power derived from FTP, and jumping metrics. The model can be expressed as follows:

$$
\begin{align}
\text{EMG}_{i} &\sim \text{Normal}(\mu_i, \sigma) \\
(1)~\mu_{i} &= \alpha + \beta_{\text{JH}} \cdot \text{JH}_{i} \\
(2)~\mu_{i} &= \alpha + \beta_{\text{FTP}} \cdot \text{FTP}_{i} \\
(3)~\mu_{i} &= \alpha + \beta_{\text{AUI}} \cdot \text{AUI}_{i} \\
(4)~\mu_{i} &= \alpha + \beta_{\text{JH}} \cdot \text{JH}_{i} + \beta_{\text{FTP}} \cdot \text{FTP}_{i} + \beta_{\text{AUI}} \cdot \text{AUI}_{i}
\end{align}
$$ {#eq-first-model}

where $\mu$ and $\sigma$ are the location and scale parameters, and $\mu$ is estimated from four linear models: 1) jump height obtained from jump tests ($\text{JH}$); 2) FTP measurements ($\text{FTP}$); 3) arm usage index obtained from jump tests ($\text{AUI}$); 4) a fully adjusted model based on the linear combination of previous predictors.

#### Muscle power and performance

In the case of muscle power indicators, our aim was to describe how performance modifying variables affect key power related metrics derived from the FTP, such as power and weight-adjusted power (WAP), which are sport-specific outcomes for cyclists. In this context, we modeled the impact of EMG and jump height, obtained from jump tests, on the magnitude of change observed in the FTP measurements in a multivariate model. The models can be expressed as:

$$
\begin{align}
\text{Power}_{i},\text{WAP}_{i} &\sim \text{Normal}(\mu_i, \sigma) \\
(1)~\mu_{i} &= \alpha + (\beta_{\text{JH}} \cdot \text{JH}_{i}) \\
(2)~\mu_{i} &= \alpha + (\beta_{\text{EMG}} \cdot \text{EMG}_{i}) \\
(3)~\mu_{i} &= \alpha + (\beta_{\text{JH}} \cdot \text{JH}_{i}) + (\beta_{\text{EMG}} \cdot \text{EMG}_{i})
\end{align}
$$ {#eq-second-model}

where $\mu$ and $\sigma$ are the location and scale parameters from the multivariate normal distribution of Power and WAP, and $\mu$ is estimated from three linear models: 1) jump height obtained from jump tests ($\text{JH}$); 2) EMG recordings ($\text{EMG}$); 3) a fully adjusted model based on the linear combination of previous predictors.

### Prior specification

For the model coefficients, we defined weakly informative priors centered at 0 as a regularization strategy during the Bayesian estimation process, shrinking potential noise-associated effects deriving from outliers or random sampling error that could arise from the observations. The intercept term's prior distribution, $\alpha$, was specified as follows:

$$
\begin{align}
\mu &= \text{Med}(\text{Y}_j) \\ 
\sigma &= \begin{cases} \text{MAD}(\text{Y}_j), & \text{if } \text{MAD}(\text{Y}_j) > 2.5 \\ 
          2.5, & \text{if } \text{MAD}(\text{Y}_j) \leq 2.5 \end{cases} \\ 
\alpha &\sim \text{Student-t}(3, \mu, \sigma)
\end{align}
$$ {#eq-alpha-prior}

where $\alpha$, specified as a t-distribution with location $\mu$ and scale $\sigma$ served as the prior distribution for the intercept, where $\mu$ represented the median ($\text{Med}$) of the response $Y$, for each variable $j$, and $\sigma$ represented the median absolute deviation (MAD) of the response variable $Y_j$, adjusted to 2.5 if the MAD was less than or equal to 2.5.

Meanwhile, the prior distributions for fixed effects $\beta$ for each variable $k$, were specified as follow:

$$
\begin{align}
\beta_k \sim \text{Normal}(0, 10)
\end{align}
$$ {#eq-b-prior}

### Model-fitting

The Bayesian generalized linear models were fitted utilizing the `brm()` function from the *brms* package [@brms], employing 5 chains with 12000 iterations per chain. The initial 2000 iterations were designated as warmup to ascertain convergence, while the remaining iterations facilitated posterior distribution estimation of model coefficients (i.e., 50000 effective iterations in total), thus offering credible intervals to substantiate parameter uncertainty.

Following the **S**equential **E**ffect e**X**istence and s**I**gnificance **T**esting (SEXIT) framework to describe the effects from Bayesian models [@makowski2019indices], the median and the CI~95%~ (using the highest density interval) were reported as a measure of centrality and uncertainty, the probability of direction (p~direction~) as measure of existence, the proportion of the posterior probability distribution that falls outside the region of practical equivalence (ROPE) as a measure of practical significance (p~signif~), estimated as one tenth (1/10 = 0.1) of the SD of the response variable, and Bayes factor (BF~10~) using Savage-Dickey density ratio against the point null indicating if the null value has become less or more likely given the observed data [@heck2019caveat], using this as a measure of absolute magnitude of evidence in favor or against the null hypothesis (of no effect).
Considering that ROPE, and therefore p~signif~, are sensitive to the predictor-level scale, continuous independent variables were standardized, which is relevant for models described in @eq-first-model and [-@eq-second-model], standardizing power and WAP as well in @eq-second-model for multivariate models.

For BF interpretation we’ve considered: BF = 1, no evidence; 1 < BF <= 3, anecdotal; 3 < BF <= 10, moderate; 10 < BF <= 30, strong; 30 < BF <= 100, very strong; and BF > 100, as extreme evidence [@jeffreys1998theory]. For the proportion of the posterior in the ROPE we considered: < 1%, significant; < 2.5%, probably significant; ≤ 97.5% & ≥ 2.5%, undecided significance; > 97.5%, probably neglible; > 99%, neglible [@makowski2019indices]. The convergence and stability of Bayesian sampling has been assessed using R-hat, which should be below 1.01 [@vehtari2019rank], and the Effective Sample Size (ESS), which should be greater than 1000 [@brms]. All the statistical analyses were computed and implemented in the R programming language [@rlanguage].

# Results

```{r paquetes}
#| include: false
if (!"pacman" %in% .packages(TRUE)) {
  warning("Package \"pacman\" is not installed.\nInstallation will be attempted...")
  install.packages("pacman")
}

pacman::p_load(data.table,
               brms,
               flextable,
               ggplot2,
               bayestestR)

source(file = here::here("R/bda/_functions.R"))

```

```{r datos, include=FALSE}
## Load the dataset
data("cyclist")
#load(here::here("docs/manuscript/misc/cyclist.RData"))

## Tables
load(here::here("docs/manuscript/tables/table-1.RData"))
load(here::here("docs/manuscript/tables/table-2.RData"))
load(here::here("docs/manuscript/tables/table-3.RData"))
```

EMG activity across different windows of measurement during FTP testing and jump performance characteristics across the different types of jump can be seen in [table 1](#tbl-1).

```{r}
#| message: false
tbl_1 |> 
  flextable::keep_with_next(part = "all")
```
[**Table 1**]{#tbl-1}. Descriptive characteristics of EMG recordings during FTP evaluation aggregated by window of measurement and summary statistics of key performance indicators of jump assessments.

## Neuromuscular efficiency

Effects derived from unadjusted models shown that arm usage index, height in Abalakov and CMJ provides anecdotal evidence in favor of the hypothesis of the presence of an effect on neuromuscular efficiency. Similar evidence was found in relation to power and WAP on median EMG. However, after estimating the marginal effect while taking into account the influence of jump and FTP paramters, we were unable to find any evidence in favour of the presence of an effect in neuromuscular efficiency (see [figure 1](#fig-1)). Univariate and adjusted model parameter effects on median EMG activity, as well as model diagnostics are shown in [table 2](#tbl-2). 

```{r}
tbl_2 |> 
  autofit(add_w = 0.01) |> 
  flextable::keep_with_next(part = "all")
```
[**Table 2**]{#tbl-2}. Model estimates and 95% CI for jump and FTP parameters associated with median EMG activity. Parameter coefficients were estimated per separate in univariate and adjusted models. 

![](figures/fig-1.pdf)
[**Figure 1**]{#fig-1}. A, Posterior distributions and shaded 50% CIs of the marginal effects of model paramters in the adjusted model; B, Empirical distribution of the response variable observed from the data to the distributions of simulated data from the posterior predictive distribution.

## Muscle power and performance

When inspecting the effect of median EMG on both power and WAP, we observed anecdotal evidence in favor of the absence of an effect and with low probability of being significant (p~signif~ ≤ 0.085) with an undecided significance based on proportion of the posterior distribution inside ROPE (Power, ROPE = 0.938; WAP, ROPE = 0.913), in the case of jump measures, we found. Univariate and adjusted model parameter effects on power and WAP, as well as model diagnostics are shown in [table 3](#tbl-3). 

```{r}
tbl_3 |> 
  autofit(add_w = 0.01) |> 
  flextable::keep_with_next(part = "all")
```
[**Table 3**]{#tbl-3}. Model estimates in terms of Z-score change in response to one SD change in independent variables and 95% CI for jump and median EMG activity parameters associated with Power and WAP. Parameter coefficients were estimated per separate in univariate and adjusted models. 

![](figures/fig-2.pdf)
[**Figure 2**]{#fig-2}. A, Posterior distributions and shaded 50% CIs of the marginal effects of model paramters in the adjusted multivariate model; B and C, Empirical distributions of the response variable observed from the data to the distributions of simulated data from the posterior predictive distribution.

# Discussion

[...].

[...].

[...].

# Conclusion

[...]. 

# Acknowledgements

[...].

# Conflict of interest

[...].

# References

<div id="refs"></div>
